---
title: "End-to-End GPS Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-End GPS Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Overview

This tutorial walks through a full `grazer` workflow:

1. Create/import GPS data.
2. Validate schema and row quality.
3. Clean obvious data issues.
4. Calculate movement and social metrics.
5. Summarise metrics by epoch.
6. Generate activity states for interpretation.

The tutorial uses a small synthetic dataset so it runs quickly and is fully reproducible.

```{r setup}
library(grazer)
```

## 1) Build an Example Dataset

In practice, most projects start with `grz_read_gps("your_file.csv")`.
Here we generate data directly to keep the tutorial self-contained.

```{r synthetic-data}
set.seed(42)

n_animals <- 3L
n_fix <- 24L * 6L * 3L # 3 days at 10-minute intervals

timestamps <- seq(
  as.POSIXct("2024-09-01 00:00:00", tz = "UTC"),
  by = "10 min",
  length.out = n_fix
)

make_track <- function(sensor_id, lon0, lat0) {
  data.frame(
    sensor_id = sensor_id,
    datetime = timestamps,
    lon = lon0 + cumsum(rnorm(n_fix, mean = 0, sd = 0.00018)),
    lat = lat0 + cumsum(rnorm(n_fix, mean = 0, sd = 0.00014)),
    stringsAsFactors = FALSE
  )
}

gps <- rbind(
  make_track("A01", 132.305, -14.474),
  make_track("A02", 132.300, -14.471),
  make_track("A03", 132.309, -14.469)
)

# Add a few realistic data issues so cleaning steps are visible.
gps <- rbind(gps, gps[1:5, ])         # duplicates
gps$lon[12] <- 0                      # invalid 0,0 coordinate
gps$lat[12] <- 0
gps$datetime[25] <- NA                # invalid datetime
gps$lon[300] <- gps$lon[300] + 0.5    # extreme jump
gps$lat[300] <- gps$lat[300] + 0.5

dim(gps)
head(gps)
```

## 2) Validate Inputs

`grz_validate()` checks required columns and row-level validity.

```{r validate}
gps_valid <- grz_validate(gps, drop_invalid = FALSE, verbose = TRUE)
```

If you want invalid rows removed immediately, use `drop_invalid = TRUE`.

## 3) Clean the GPS Data

`grz_clean()` applies selected cleaning steps in sequence.

```{r clean}
gps_clean <- grz_clean(
  gps_valid,
  steps = c("duplicates", "errors", "speed_fixed", "denoise"),
  max_speed_mps = 4,
  verbose = TRUE
)

dim(gps_clean)
```

## 4) Calculate Row-Level Metrics

Movement metrics are appended per row:

```{r movement}
gps_move <- grz_calculate_movement(gps_clean, verbose = FALSE)

head(gps_move[, c("sensor_id", "datetime", "step_m", "speed_mps", "turn_rad")])
```

Social metrics are calculated from pairwise distances at matching timestamps:

```{r social}
gps_social <- grz_calculate_social(
  gps_clean,
  interpolate = FALSE,
  verbose = FALSE
)

head(gps_social[, c("sensor_id", "datetime", "nearest_neighbor_m", "mean_dist_to_others_m")])
```

## 5) Summarise by Epoch

Create daily summaries across movement, social, and spatial metric blocks:

```{r epoch}
daily_metrics <- grz_calculate_epoch_metrics(
  gps_clean,
  epoch = "day",
  include = c("movement", "social", "spatial"),
  verbose = FALSE
)

head(daily_metrics)
```

## 6) Activity State Interpretation

`grz_classify_activity_consensus()` combines HMM and spatial signals into a final
`active`/`inactive` state.

```{r behavior}
gps_states <- grz_classify_activity_consensus(
  gps_clean,
  groups = "sensor_id",
  verbose = FALSE
)

table(gps_states$activity_state_consensus, useNA = "ifany")
```

For additional diagnostics and threshold tuning, see:

- `grz_plot_diurnal_metrics()`
- `grz_behavior_threshold_guide()`
- `grz_tune_thresholds()`
- `grz_validate_behavior()`

## 7) Interactive QA and Manual Labelling

The two interactive tools are:

- `grz_map()` for exploratory mapping.
- `grz_label_gps_states()` for manual `ACTIVE`/`INACTIVE` labels.

Use these in an interactive R session:

```{r interactive-tools, eval = FALSE}
# Interactive map
grz_map(
  data = gps_states,
  timeline = TRUE,
  group = "sensor_id",
  state_col = "activity_state_consensus"
)

# Interactive manual state labelling app
labelled <- grz_label_gps_states(
  data = gps_states,
  lon = "lon",
  lat = "lat",
  time = "datetime",
  color_by = "sensor_id",
  initial_label_col = "label"
)
```

## 8) Save Outputs

```{r save}
# Example export pattern
# data.table::fwrite(daily_metrics, "daily_metrics.csv")
# data.table::fwrite(gps_states, "gps_states.csv")
```

## Next Steps

- Replace synthetic data with your project CSV files.
- Add paddock polygons with `grz_clean_spatial()` and `grz_append_paddock_names()`.
- Use `grz_label_gps_states()` labels to benchmark and tune consensus settings.
