---
title: "GPS 201: Activity Classification and Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPS 201: Activity Classification and Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Overview

This tutorial covers the next stage after cleaning and metric generation:

1. classify each GPS row as `active` or `inactive`,
2. manually label rows in an interactive app,
3. compare model predictions against labels,
4. rank parameter settings by agreement.

## Why This Workflow?

State classification is iterative. A clear validation loop gives better decisions:

- You can inspect how model assumptions map to real movement behaviour.
- You can use manual labels as a transparent ground-truth reference.
- You can tune parameters with a reproducible and auditable process.
- You can report uncertainty and model agreement directly.

## 1) Load packages

```{r}
library(grazer)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```

## 2) Build a small reproducible example

```{r}
set.seed(202)

timestamps <- seq(
  from = as.POSIXct("2024-06-01 00:00:00", tz = "UTC"),
  by = "10 min",
  length.out = 3 * 24 * 6
)

animal_info <- tibble(
  sensor_id = c("A01", "A02"),
  lon0 = c(132.310, 132.316),
  lat0 = c(-14.462, -14.467)
)

gps_list <- list()
for (i in seq_len(nrow(animal_info))) {
  gps_list[[i]] <- tibble(
    sensor_id = animal_info$sensor_id[i],
    datetime = timestamps,
    lon = animal_info$lon0[i] + cumsum(rnorm(length(timestamps), 0, 0.00020)),
    lat = animal_info$lat0[i] + cumsum(rnorm(length(timestamps), 0, 0.00016))
  )
}
gps <- bind_rows(gps_list)

gps <- grz_clean(
  data = gps,
  steps = c("errors", "speed_fixed", "denoise"),
  max_speed_mps = 4,
  verbose = FALSE
)
```

How this works in practice:

- Start from already cleaned GPS rows with stable identifiers.
- Keep example windows small enough to iterate quickly while tuning.

Common pitfalls and checks:

- Pitfall: tuning on too broad a window early on.  
  Check: begin with one animal and a short period, then scale up.

## 3) Run consensus active/inactive classification

`grz_classify_activity_consensus()` combines HMM and spatial signals into one
final state column.

```{r}
gps_states <- grz_classify_activity_consensus(
  data = gps,
  groups = "sensor_id",
  verbose = FALSE
)

state_view <- gps_states %>%
  as_tibble() %>%
  select(sensor_id, datetime, lon, lat, activity_state_consensus)

state_view

state_more <- setdiff(names(gps_states), names(state_view))
cat(
  "i",
  length(state_more),
  "more columns:",
  paste(head(state_more, 10), collapse = ", "),
  if (length(state_more) > 10) ", ..." else "",
  "\n"
)
```

Key state variables in this workflow:

- `activity_state_hmm`: HMM-based `active`/`inactive` estimate.
- `inactive_prob_hmm`: HMM inactivity probability.
- `activity_state_spatial`: spatial clustering based estimate.
- `activity_state_consensus`: final state after combining methods.
- `inactive_score_consensus`: consensus inactivity score (0 to 1).

How this works in practice:

- HMM uses movement features to estimate latent activity states.
- Spatial classifier identifies local dwell/stay behaviour from position patterns.
- Consensus combines both signals into a final `active`/`inactive` decision.

Common pitfalls and checks:

- Pitfall: assuming one classifier is always correct.  
  Check: inspect both component outputs (`hmm` and `spatial`) before consensus.
- Pitfall: class imbalance leading to misleading accuracy.  
  Check: inspect class proportions by day and animal.

```{r}
gps_states %>%
  as_tibble() %>%
  ggplot(aes(x = lon, y = lat, color = activity_state_consensus)) +
  geom_point(alpha = 0.7, size = 1.3) +
  facet_wrap(~sensor_id, scales = "free") +
  scale_color_manual(values = c(active = "#1a9641", inactive = "#d7191c")) +
  labs(
    title = "State Classification in Space",
    x = "Longitude",
    y = "Latitude",
    color = "Consensus state"
  ) +
  theme_minimal()
```

```{r}
grz_plot_diurnal_states(
  data = gps_states,
  state_col = "activity_state_consensus",
  group_col = "sensor_id",
  plot_type = "line"
)
```

```{r}
gps_states %>%
  as_tibble() %>%
  mutate(day = as.Date(datetime)) %>%
  group_by(sensor_id, day, activity_state_consensus) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(sensor_id, day) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = day, y = prop, fill = activity_state_consensus)) +
  geom_col(position = "stack") +
  facet_wrap(~sensor_id) +
  scale_fill_manual(values = c(active = "#1a9641", inactive = "#d7191c")) +
  labs(
    title = "Daily Proportion of Active vs Inactive States",
    x = "Day",
    y = "Proportion of fixes",
    fill = "Consensus state"
  ) +
  theme_minimal()
```

## 4) Add manual labels (interactive app)

Use this in an interactive R session to create or edit a `label` column.

```{r, eval = FALSE}
gps_states$point_id <- paste(
  gps_states$sensor_id,
  format(gps_states$datetime, "%Y%m%d%H%M%S", tz = "UTC"),
  seq_len(nrow(gps_states)),
  sep = "_"
)

labelled <- grz_label_gps_states(
  data = gps_states,
  lon = "lon",
  lat = "lat",
  time = "datetime",
  id = "point_id",
  color_by = "sensor_id",
  initial_label_col = "label",
  start_day_offset = 0L,
  time_window = "week",
  n_animals = 1L,
  animal_col = "sensor_id"
)
```

If you save that output to CSV, you can reload it later and continue editing:

```{r, eval = FALSE}
# data.table::fwrite(labelled, "labelled_states.csv")
# labelled <- data.table::fread("labelled_states.csv")
# labelled$datetime <- as.POSIXct(labelled$datetime, tz = "UTC")
```

How this works in practice:

- Manual labels provide ground truth for calibration and validation.
- Labels should be saved with stable IDs so they can be reused across sessions.

Common pitfalls and checks:

- Pitfall: relabelling the same points without versioning.  
  Check: keep dated label files or append a reviewer/version column.
- Pitfall: using ambiguous behaviour windows.  
  Check: mark uncertain points as `NA` instead of forcing a label.

## 5) Demonstrate a prediction-vs-label comparison

For a fully reproducible vignette, we create a synthetic `label` column by
copying predictions and flipping a random subset of rows.  
In your project, replace this with manual labels from `grz_label_gps_states()`.

```{r}
labelled <- gps_states
labelled$label <- labelled$activity_state_consensus

set.seed(303)
flip_n <- floor(0.1 * nrow(labelled))
flip_idx <- sample(seq_len(nrow(labelled)), size = flip_n)

labelled$label[flip_idx] <- ifelse(
  labelled$label[flip_idx] == "active",
  "inactive",
  "active"
)

label_view <- labelled %>%
  as_tibble() %>%
  select(sensor_id, datetime, activity_state_consensus, label, inactive_score_consensus)

label_view

label_more <- setdiff(names(labelled), names(label_view))
cat(
  "i",
  length(label_more),
  "more columns:",
  paste(head(label_more, 10), collapse = ", "),
  if (length(label_more) > 10) ", ..." else "",
  "\n"
)
```

How this works in practice:

- Truth labels and predicted states are compared on the same rows.
- Only valid binary states (`active`, `inactive`) are included for scoring.

Common pitfalls and checks:

- Pitfall: comparing with stale predictions from different preprocessing.  
  Check: regenerate predictions from the same cleaned dataset used for labels.
- Pitfall: hidden leakage from copying predicted labels into truth labels.  
  Check: keep manually-labelled files separate from model outputs.

## 6) Tune parameters and rank the top 3 settings

```{r}
grid <- tidyr::expand_grid(
  inactive_threshold = c(0.50, 0.55, 0.60),
  spatial_radius_m = c(15, 20),
  spatial_min_dwell_mins = c(10, 20)
)

results <- tibble()

for (i in seq_len(nrow(grid))) {
  pred <- grz_classify_activity_consensus(
    data = labelled,
    groups = "sensor_id",
    decision_rule = "weighted",
    inactive_threshold = grid$inactive_threshold[[i]],
    spatial_radius_m = grid$spatial_radius_m[[i]],
    spatial_min_dwell_mins = grid$spatial_min_dwell_mins[[i]],
    spatial_min_points = 3L,
    min_run_n = 2L,
    verbose = FALSE
  )

  comparison <- tibble(
    truth = tolower(trimws(as.character(labelled$label))),
    pred = tolower(trimws(as.character(pred$activity_state_consensus)))
  ) %>%
    filter(truth %in% c("active", "inactive")) %>%
    filter(pred %in% c("active", "inactive"))

  n_compared <- nrow(comparison)
  accuracy <- if (n_compared == 0L) {
    NA_real_
  } else {
    mean(comparison$truth == comparison$pred)
  }

  results <- bind_rows(
    results,
    tibble(
      inactive_threshold = grid$inactive_threshold[[i]],
      spatial_radius_m = grid$spatial_radius_m[[i]],
      spatial_min_dwell_mins = grid$spatial_min_dwell_mins[[i]],
      n_compared = n_compared,
      accuracy = accuracy
    )
  )
}

top3 <- results %>%
  arrange(desc(accuracy), desc(n_compared)) %>%
  slice_head(n = 3) %>%
  mutate(accuracy_percent = round(100 * accuracy, 2)) %>%
  select(inactive_threshold, spatial_radius_m, spatial_min_dwell_mins, n_compared, accuracy_percent)

top3
```

How this works in practice:

- A parameter grid is evaluated with a consistent consensus rule.
- Each setting is ranked by observed agreement against manual labels.
- Top settings are candidates for holdout validation, not automatic final choices.

Common pitfalls and checks:

- Pitfall: overfitting thresholds to a single week/animal.  
  Check: validate top settings on held-out animals and periods.
- Pitfall: choosing accuracy only.  
  Check: also inspect confusion matrix and error type balance.

```{r}
results %>%
  mutate(accuracy_percent = 100 * accuracy) %>%
  ggplot(aes(x = factor(inactive_threshold), y = factor(spatial_radius_m), fill = accuracy_percent)) +
  geom_tile(color = "white") +
  facet_wrap(~spatial_min_dwell_mins) +
  labs(
    title = "Consensus Accuracy Across Parameter Grid",
    x = "inactive_threshold",
    y = "spatial_radius_m",
    fill = "Accuracy (%)"
  ) +
  theme_minimal()
```

```{r}
best <- top3 %>% slice(1)

best_pred <- grz_classify_activity_consensus(
  data = labelled,
  groups = "sensor_id",
  decision_rule = "weighted",
  inactive_threshold = best$inactive_threshold[[1]],
  spatial_radius_m = best$spatial_radius_m[[1]],
  spatial_min_dwell_mins = best$spatial_min_dwell_mins[[1]],
  spatial_min_points = 3L,
  min_run_n = 2L,
  verbose = FALSE
)

confusion_df <- tibble(
  truth = tolower(trimws(as.character(labelled$label))),
  pred = tolower(trimws(as.character(best_pred$activity_state_consensus)))
) %>%
  filter(truth %in% c("active", "inactive")) %>%
  filter(pred %in% c("active", "inactive")) %>%
  count(truth, pred, .drop = FALSE)

confusion_df %>%
  ggplot(aes(x = truth, y = pred, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 5) +
  labs(
    title = "Confusion Matrix for Best Parameter Set",
    x = "Manual label",
    y = "Predicted state",
    fill = "Count"
  ) +
  theme_minimal()
```

## 7) Optional advanced diagnostics

These helpers are useful once you have enough labelled data:

```{r, eval = FALSE}
# Diurnal feature patterns for threshold setting
grz_plot_diurnal_metrics(
  data = labelled,
  metrics = c("step_m", "turn_rad"),
  group_col = "sensor_id"
)

# Behaviour diagnostics (feature summary, transitions, bouts, PCA)
diagnostics <- grz_validate_behavior(
  data = labelled,
  state_col = "activity_state_consensus",
  groups = "sensor_id",
  pca = TRUE
)
```

`grz_validate_behavior()` returns state counts, transitions, bout summaries, and
an optional PCA diagnostic for the selected state column.

How this works in practice:

- Transition and bout outputs help detect unstable state switching.
- PCA and feature summaries help assess separation between labelled states.

Common pitfalls and checks:

- Pitfall: interpreting PCA clusters as definitive behaviour classes.  
  Check: use PCA as support, not as the sole decision criterion.

## Recommended Practice

1. Label a representative subset (for example, one animal for one week).
2. Tune parameters on that subset.
3. Validate performance metrics.
4. Freeze parameters and run full-cohort processing.
5. Re-check performance when deployment conditions change.

## References

- Zucchini, W., MacDonald, I. L., & Langrock, R. (2016).  
  *Hidden Markov Models for Time Series: An Introduction Using R* (2nd ed.).  
  Core reference for HMM-based state modelling.
- Li, Q., Zheng, Y., Xie, X., Chen, Y., Liu, W., & Ma, W.-Y. (2008).  
  *Mining user similarity based on location history*. ACM SIGSPATIAL GIS.  
  Includes stay-point style concepts used in spatial inactivity detection.
