---
title: "GPS 101: Core Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPS 101: Core Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Overview

This tutorial walks through the main `grazer` workflow for cattle GPS data:

1. Start from GPS rows in canonical columns (`sensor_id`, `datetime`, `lon`, `lat`).
2. Validate required schema and data types.
3. Run an interactive playback check to inspect trajectories over time.
4. Clean obvious errors and noise.
5. Calculate movement, social, and spatial metrics.
6. Summarise results to daily metrics for interpretation.

## Why This Workflow?

A structured pipeline improves reproducibility and interpretation:

- You detect data issues before model fitting.
- You apply cleaning rules consistently across experiments.
- You generate comparable metrics across paddocks, cohorts, and deployments.
- You keep row-level and day-level outputs aligned for QA and reporting.

## 1) Load packages

```{r}
library(grazer)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```

## 2) Create an example GPS table

This vignette uses synthetic data so it is fully reproducible.  
In real projects, start by reading your file, mapping to the required schema,
then validating:

```{r, eval = FALSE}
gps_raw <- readr::read_csv("your_file.csv", show_col_types = FALSE)

# Required canonical columns for grazer:
# sensor_id, datetime, lon, lat
gps <- dplyr::rename(
  gps_raw,
  sensor_id = device_id,
  datetime = Time,
  lon = Longitude,
  lat = Latitude
)

gps <- grz_validate(gps, drop_invalid = FALSE, verbose = TRUE)
```

How this works in practice:

- Keep raw ingest as close as possible to source files.
- Add only minimal transformations before validation.
- Preserve metadata columns so they can be used later in grouping and summaries.

Common pitfalls and checks:

- Pitfall: parsing datetimes in local time by accident.  
  Check: confirm all times are parsed in the same timezone (typically UTC).
- Pitfall: silently coercing text coordinates to `NA`.  
  Check: inspect rows with missing/invalid `lon`/`lat` before cleaning.

```{r}
set.seed(101)

timestamps <- seq(
  from = as.POSIXct("2024-05-01 00:00:00", tz = "UTC"),
  by = "10 min",
  length.out = 3 * 24 * 6
)

animal_info <- tibble(
  sensor_id = c("A01", "A02", "A03"),
  lon0 = c(132.305, 132.299, 132.311),
  lat0 = c(-14.474, -14.471, -14.468)
)

gps_list <- list()
for (i in seq_len(nrow(animal_info))) {
  gps_list[[i]] <- tibble(
    sensor_id = animal_info$sensor_id[i],
    datetime = timestamps,
    lon = animal_info$lon0[i] + cumsum(rnorm(length(timestamps), 0, 0.00018)),
    lat = animal_info$lat0[i] + cumsum(rnorm(length(timestamps), 0, 0.00015))
  )
}
gps_raw <- bind_rows(gps_list)

# Add a few realistic issues so validation and cleaning have visible effects.
gps_raw <- bind_rows(gps_raw, gps_raw %>% slice(1:5)) # duplicate rows
gps_raw$lon[20] <- 0                                  # invalid location (0,0)
gps_raw$lat[20] <- 0
gps_raw$datetime[33] <- NA                            # invalid datetime

nrow(gps_raw)
head(gps_raw)
```

## 3) Validate required columns and row values

```{r}
gps_valid <- grz_validate(
  data = gps_raw,
  drop_invalid = FALSE,
  verbose = TRUE
)

validation_qc <- attr(gps_valid, "validation_qc")
invalid_rows <- attr(gps_valid, "invalid_rows")

validation_qc
head(invalid_rows)
```

The validation object provides:

- `validation_qc`: counts/proportions of invalid records by issue type.
- `invalid_rows`: full rows that failed validation checks, with `invalid_reason`.

How this works in practice:

- `grz_validate()` enforces the minimum schema (`sensor_id`, `datetime`, `lon`, `lat`).
- It parses and type-checks each core field.
- It records invalid reasons at row level so decisions are auditable.

Common pitfalls and checks:

- Pitfall: dropping invalid rows too early can hide data-logger issues.  
  Check: run once with `drop_invalid = FALSE` and inspect `invalid_rows`.
- Pitfall: blank `sensor_id` values across multiple animals.  
  Check: verify `sensor_id` uniqueness and completeness before movement metrics.

## 4) Interactive playback quick look (after validate)

Before cleaning and feature derivation, it is useful to visually inspect the
validated trajectories in a timeline map.

```{r}
gps_valid_map <- grz_validate(
  data = gps_raw,
  drop_invalid = TRUE,
  verbose = FALSE
)

playback_map <- grz_playback_gps(
  data = gps_valid_map,
  group = "sensor_id",
  align = TRUE,
  align_interval_mins = 20,
  tail_points = 20,
  show_points = TRUE,
  point_size_slider = TRUE,
  point_size_min = 1,
  point_size_max = 20,
  playback_steps = 800,
  playback_duration_ms = 18000,
  popup_fields = c("sensor_id", "datetime"),
  warnings = FALSE,
  progress = FALSE,
  show_loading_overlay = TRUE
)

playback_map
```

How this works in practice:

- Playback gives an immediate quality check for track continuity and obvious artefacts.
- The timeline control helps inspect movement over time before committing to cleaning settings.
- Point-size slider makes sparse and dense periods easier to inspect.

## 5) Clean the GPS rows

This run applies four common cleaning steps:

- `duplicates`: removes repeated fixes.
- `errors`: drops invalid coordinates and malformed rows.
- `speed_fixed`: removes biologically implausible moves.
- `denoise`: reduces local GPS bounce in static periods.

```{r}
gps_clean <- grz_clean(
  data = gps_valid,
  steps = c("duplicates", "errors", "speed_fixed", "denoise"),
  max_speed_mps = 4,
  verbose = TRUE
)

nrow(gps_clean)
head(gps_clean)
```

How this works in practice:

- Cleaning is intentionally drop-based and ordered.
- Duplicate and invalid coordinate fixes are removed first.
- Speed filtering removes biologically implausible jumps.
- Denoising removes jitter during likely static periods.

Common pitfalls and checks:

- Pitfall: speed threshold too strict for your production class (e.g., mustering).  
  Check: inspect upper speed tail before setting `max_speed_mps`.
- Pitfall: aggressive denoising in coarse sampling intervals.  
  Check: compare pre/post row counts by animal and day.

## 6) Calculate movement metrics (row-level)

```{r}
gps_movement <- grz_calculate_movement(gps_clean, verbose = FALSE)

movement_view <- gps_movement %>%
  as_tibble() %>%
  select(sensor_id, datetime, step_m, speed_mps, turn_rad)

movement_view

movement_more <- setdiff(names(gps_movement), names(movement_view))
cat(
  "i",
  length(movement_more),
  "more columns:",
  paste(head(movement_more, 10), collapse = ", "),
  if (length(movement_more) > 10) ", ..." else "",
  "\n"
)
```

Key movement variables:

- `step_m`: great-circle distance between consecutive fixes.
- `speed_mps`: step distance divided by elapsed time in seconds.
- `turn_rad`: absolute turning angle (radians) between consecutive bearings.

How this works in practice:

- `step_m` is great-circle distance between consecutive fixes.
- `speed_mps` is `step_m / step_dt_s`.
- `turn_rad` uses the absolute difference in successive headings.

Common pitfalls and checks:

- Pitfall: irregular fix intervals inflate/deflate apparent speed.  
  Check: inspect `step_dt_s` distribution.
- Pitfall: first fix per track has no lag values.  
  Check: expect `NA` in first row of each group for lag-derived metrics.

```{r}
gps_movement %>%
  as_tibble() %>%
  ggplot(aes(x = step_m, fill = sensor_id)) +
  geom_histogram(bins = 40, alpha = 0.55, position = "identity") +
  labs(
    title = "Step Distance Distribution",
    x = "Step distance (m)",
    y = "Count",
    fill = "Sensor ID"
  ) +
  theme_minimal()
```

```{r}
gps_movement %>%
  as_tibble() %>%
  mutate(hour = as.integer(format(datetime, "%H", tz = "UTC"))) %>%
  ggplot(aes(x = hour, y = speed_mps, color = sensor_id)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(se = FALSE, linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 23, by = 3)) +
  labs(
    title = "Diurnal Speed Pattern",
    x = "Hour of day (UTC)",
    y = "Speed (m/s)",
    color = "Sensor ID"
  ) +
  theme_minimal()
```

## 7) Calculate social metrics (row-level)

```{r}
gps_social <- grz_calculate_social(
  gps_movement,
  thresholds_m = c(30, 50, 100),
  interpolate = FALSE,
  verbose = FALSE
)

n30_col <- grep("^n_within_\\s*30m$", names(gps_social), value = TRUE)
if (length(n30_col) == 0L) {
  stop("Expected a 30 m neighbour count column after grz_calculate_social().")
}

gps_social_tbl <- gps_social %>%
  as_tibble() %>%
  mutate(n_within_30m = .data[[n30_col[[1]]]])

social_view <- gps_social_tbl %>%
  select(sensor_id, datetime, nearest_neighbor_m, n_within_30m, mean_dist_to_others_m)

social_view

social_more <- setdiff(names(gps_social), names(social_view))
cat(
  "i",
  length(social_more),
  "more columns:",
  paste(head(social_more, 10), collapse = ", "),
  if (length(social_more) > 10) ", ..." else "",
  "\n"
)
```

Key social variables:

- `nearest_neighbor_m`: distance to closest other animal at matching timestamp.
- `n_within_30m`: number of other animals within 30 m.
- `mean_dist_to_others_m`: average distance to all other animals.

How this works in practice:

- For each timestamp (and optional herd partition), pairwise distances are computed.
- Nearest-neighbour metrics and threshold counts are derived from the distance matrix.
- Threshold columns (`n_within_*m`, `any_within_*m`) support density/proximity interpretation.

Common pitfalls and checks:

- Pitfall: unsynchronised timestamps between collars.  
  Check: use `grz_align()` when needed before social metrics.
- Pitfall: interpreting social isolation in very small group sizes.  
  Check: inspect `social_group_size` before interpretation.

```{r}
gps_social_tbl %>%
  ggplot(aes(x = nearest_neighbor_m, fill = sensor_id)) +
  geom_density(alpha = 0.35) +
  labs(
    title = "Nearest-Neighbour Distance Density",
    x = "Nearest neighbour distance (m)",
    y = "Density",
    fill = "Sensor ID"
  ) +
  theme_minimal()
```

```{r}
gps_social_tbl %>%
  mutate(day = as.Date(datetime)) %>%
  group_by(sensor_id, day) %>%
  summarise(mean_n_within_30m = mean(n_within_30m, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = day, y = mean_n_within_30m, color = sensor_id)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  labs(
    title = "Daily Mean Number of Animals Within 30 m",
    x = "Day",
    y = "Mean n_within_30m",
    color = "Sensor ID"
  ) +
  theme_minimal()
```

## 8) Summarise to day-level movement, social, and spatial outputs

```{r}
daily_metrics <- grz_calculate_epoch_metrics(
  data = gps_social,
  epoch = "day",
  include = c("movement", "social", "spatial"),
  verbose = FALSE
)

summary_view <- daily_metrics %>%
  as_tibble() %>%
  select(sensor_id, epoch, total_distance_m, mean_nn_m, mcp95_area_ha)

summary_view

summary_more <- setdiff(names(daily_metrics), names(summary_view))
cat(
  "i",
  length(summary_more),
  "more columns:",
  paste(head(summary_more, 10), collapse = ", "),
  if (length(summary_more) > 10) ", ..." else "",
  "\n"
)
```

Key summary variables:

- `total_distance_m`: total distance moved over the epoch.
- `mean_nn_m`: mean nearest-neighbour distance over the epoch.
- `mcp95_area_ha`: 95% minimum convex polygon home-range area (ha).

How this works in practice:

- Epoch summaries aggregate row-level metrics to `day`, `week`, or `month`.
- Movement, social, and spatial blocks are merged on shared keys.
- Spatial area metrics are only meaningful with adequate fixes per epoch.

Common pitfalls and checks:

- Pitfall: over-interpreting area estimates from sparse days.  
  Check: screen by `n_fixes` before comparing `mcp95_area_ha`.
- Pitfall: combining epochs with different grazing contexts.  
  Check: facet/stratify by paddock, cohort, or treatment when available.

```{r}
daily_metrics %>%
  as_tibble() %>%
  mutate(epoch = as.Date(epoch)) %>%
  ggplot(aes(x = epoch, y = total_distance_m, color = sensor_id, group = sensor_id)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  labs(
    title = "Daily Total Distance by Animal",
    x = "Day",
    y = "Total distance (m)",
    color = "Sensor ID"
  ) +
  theme_minimal()
```

```{r}
daily_metrics %>%
  as_tibble() %>%
  mutate(epoch = as.Date(epoch)) %>%
  ggplot(aes(x = epoch, y = mcp95_area_ha, color = sensor_id, group = sensor_id)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  labs(
    title = "Daily 95% MCP Area by Animal",
    x = "Day",
    y = "MCP95 area (ha)",
    color = "Sensor ID"
  ) +
  theme_minimal()
```

## 9) Diurnal feature plots

`grz_plot_diurnal_metrics()` gives an interpretable view of how movement
features vary by hour-of-day.

How this works in practice:

- Hour-of-day aggregation is applied per metric and group.
- Heatmaps highlight temporal structure that is useful for threshold setting.

Common pitfalls and checks:

- Pitfall: timezone mismatch between sensor logging and ecological interpretation.  
  Check: set and verify timezone assumptions before diurnal plotting.

```{r}
grz_plot_diurnal_metrics(
  data = gps_social,
  metrics = c("step_m", "turn_rad"),
  group_col = "sensor_id",
  agg_fun = "median",
  scale = "none"
)
```

## Next Step

Continue to the 201 tutorial for:

- active/inactive classification,
- manual state labelling with `grz_label_gps_states()`,
- and prediction-vs-label validation.

## References

- Sinnott, R. W. (1984). *Virtues of the Haversine*. Sky and Telescope, 68(2), 159.  
  Distance calculations in `grazer` use the great-circle (haversine) approach.
- Burt, W. H. (1943). *Territoriality and home range concepts as applied to mammals*.  
  Journal of Mammalogy, 24(3), 346-352.  
  Conceptual basis for home-range summaries such as MCP.
