---
title: "GPS 101: Core Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPS 101: Core Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This tutorial is a practical introduction to the `grazer` core pipeline:

1. Start from GPS rows in canonical columns (`sensor_id`, `datetime`, `lon`, `lat`).
2. Validate required schema and data types.
3. Clean obvious errors and noise.
4. Calculate movement, social, and spatial metrics.
5. Summarise results to daily metrics for interpretation.

## Why This Workflow?

A structured workflow helps avoid common analysis problems:

- You detect data issues before model fitting.
- You apply cleaning rules consistently across experiments.
- You generate comparable metrics across paddocks, cohorts, and deployments.
- You keep row-level and day-level outputs aligned for QA and reporting.

## 1) Load the package

```{r}
library(grazer)
```

## 2) Create an example GPS table

This vignette builds synthetic data so it is fully reproducible.  
In real projects you would typically call `grz_read_gps("your_file.csv")`.

```{r}
set.seed(101)

timestamps <- seq(
  from = as.POSIXct("2024-05-01 00:00:00", tz = "UTC"),
  by = "10 min",
  length.out = 3 * 24 * 6
)

animal_ids <- c("A01", "A02", "A03")
gps_raw <- data.frame()

for (i in seq_along(animal_ids)) {
  one_animal <- data.frame(
    sensor_id = animal_ids[i],
    datetime = timestamps,
    lon = 132.30 + (i * 0.004) + cumsum(rnorm(length(timestamps), 0, 0.00018)),
    lat = -14.47 - (i * 0.002) + cumsum(rnorm(length(timestamps), 0, 0.00015)),
    stringsAsFactors = FALSE
  )
  gps_raw <- rbind(gps_raw, one_animal)
}

# Add a few realistic issues so validation and cleaning have visible effects.
gps_raw <- rbind(gps_raw, gps_raw[1:5, ])   # duplicate rows
gps_raw$lon[20] <- 0                         # invalid location (0,0)
gps_raw$lat[20] <- 0
gps_raw$datetime[33] <- NA                   # invalid datetime

nrow(gps_raw)
head(gps_raw)
```

## 3) Validate required columns and row values

```{r}
gps_valid <- grz_validate(
  data = gps_raw,
  drop_invalid = FALSE,
  verbose = TRUE
)

validation_qc <- attr(gps_valid, "validation_qc")
invalid_rows <- attr(gps_valid, "invalid_rows")

validation_qc
head(invalid_rows)
```

## 4) Clean the GPS rows

This run applies four common cleaning steps:

- `duplicates`: removes repeated fixes.
- `errors`: drops invalid coordinates and malformed rows.
- `speed_fixed`: removes biologically implausible moves.
- `denoise`: reduces local GPS bounce in static periods.

```{r}
gps_clean <- grz_clean(
  data = gps_valid,
  steps = c("duplicates", "errors", "speed_fixed", "denoise"),
  max_speed_mps = 4,
  verbose = TRUE
)

nrow(gps_clean)
head(gps_clean)
```

## 5) Calculate row-level movement and social metrics

```{r}
gps_movement <- grz_calculate_movement(gps_clean, verbose = FALSE)
gps_metrics <- grz_calculate_social(gps_movement, interpolate = FALSE, verbose = FALSE)

head(gps_metrics[, c(
  "sensor_id", "datetime", "step_m", "speed_mps", "turn_rad",
  "nearest_neighbor_m", "mean_dist_to_others_m"
)])
```

## 6) Summarise to day-level movement, social, and spatial outputs

```{r}
daily_metrics <- grz_calculate_epoch_metrics(
  data = gps_metrics,
  epoch = "day",
  include = c("movement", "social", "spatial"),
  verbose = FALSE
)

daily_metrics
```

## 7) Plot key features for quick QA

Use quick plots to check whether values are biologically reasonable.

```{r}
par(mfrow = c(1, 2))

hist(
  gps_metrics$step_m,
  breaks = 40,
  main = "Step Distance (m)",
  xlab = "step_m",
  col = "grey80",
  border = "white"
)

hist(
  gps_metrics$speed_mps,
  breaks = 40,
  main = "Speed (m/s)",
  xlab = "speed_mps",
  col = "grey80",
  border = "white"
)

par(mfrow = c(1, 1))
```

## Next Step

Continue to the 201 tutorial for:

- active/inactive classification,
- manual state labelling with `grz_label_gps_states()`,
- and prediction-vs-label validation.
