---
title: "GPS 101: Core Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPS 101: Core Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## Overview

This tutorial is written for animal scientists who are new to GPS workflows.
It uses the Manbulloo dataset and follows a practical, investigator-led process:

1. Bring in the data.
2. Validate schema and minimum requirements.
3. Clean the data with individual cleaning functions (not the wrapper).
4. Calculate movement metrics.
5. Summarise movement and inspect diurnal structure.
6. Calculate social metrics.
7. Summarise social metrics.
8. Summarise spatial metrics and week-to-week overlap.
9. Show helper visualisation functions and conclude.

## 0) Load packages

```{r}
library(grazer)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

has_sf <- requireNamespace("sf", quietly = TRUE)
if (has_sf) {
  library(sf)
}
```

## 1) Bring in the data

For this tutorial we use the in-built dataset. This is simulated dataset, trained off real gps tracking data from for diverse regions. The paddock used for the simulation is a random paddock identfied and mapped via satellite imagery for the purposes of simulating a livestock tracking dataset. The simulated cattle presented in this tutorial have no associated with the land owner.

To keep runtime practical while still being representative, we subset to:
- first 15 collars
- first 12 weeks from deployment start


```{r}
data_file <- file.path("test_data", "ManbullooToSimulate.csv")

gps_raw <- read.csv(data_file)

gps_raw <- gps_raw |> mutate(datetime = as.POSIXct(datetime))

sensor_keep <- sort(unique(gps_raw$sensor_id))
sensor_keep <- sensor_keep[seq_len(min(15L, length(sensor_keep)))]

start_time <- min(gps_raw$datetime, na.rm = TRUE)
end_time <- start_time + as.difftime(12L, units = "weeks")

gps <- gps_raw %>%
    filter(sensor_id %in% sensor_keep,
     datetime >= start_time, datetime < end_time)

dplyr::glimpse(gps)
```

### 1b) Quick interactive map preview (first 4 days)

One for the first things to do when we're using GPS data is to actually look at it. While this sounds basic, you will get to a point of proficiency where you start to skip this step. But, let me assure you, at some point in time you will get caught out if you don't do this step! Grazer has a html-based interactive mapping feature which will render a leaflet map with a timeline to allow you to scrub through your data and zoom in on areas of interest. It will take some time to render. Be patient. Reducing the time period, number of animals, or plotting every nth point will decrease render time.

```{r, eval = FALSE}
map_4d <- gps %>%
  filter(datetime < min(datetime, na.rm = TRUE) + as.difftime(4, units = "days"))

grz_map(
  data = map_4d,
  group = "sensor_id",
  timeline = TRUE,
  popup_fields = c("sensor_id", "datetime"),
  warnings = FALSE
)
```

## 2) Validate and explain minimum schema

`grazer` minimum required columns:
- `sensor_id`: collar/animal ID
- `datetime`: timestamp
- `lon`: longitude (WGS84)
- `lat`: latitude (WGS84)

Manbulloo is already in that schema, so we can validate directly.

```{r}
gps_valid_report <- grz_validate(gps, drop_invalid = FALSE, verbose = FALSE)
val_qc <- attr(gps_valid_report, "validation_qc")
invalid_rows <- attr(gps_valid_report, "invalid_rows")

val_qc
head(invalid_rows)
```

For cleaning we usually proceed with invalid rows dropped:

```{r}
gps_valid <- grz_validate(gps, drop_invalid = TRUE, verbose = FALSE)
```

## 3) Start cleaning (individual functions first)

### 3a) Remove duplicates

```{r}
gps_dedup <- grz_clean_duplicates(gps_valid, verbose = FALSE)
```

### 3b) Filter to paddock extent (100 m buffer)

Put your paddock boundary KML in `data/paddock.kml`.
KML imports commonly expose paddock names in `Name` or `Description`;
`grz_clean_spatial()` accepts either.

```{r}
paddock_path <- file.path("test_data", "paddock.kml")

if (has_sf) {
  paddocks_sf <- sf::st_read(paddock_path, quiet = TRUE)

  gps_spatial <- grz_clean_spatial(
      data = gps_dedup,
      paddocks_sf = paddocks_sf,
      buffer_m = 100,
      append_paddock = TRUE,
      verbose = FALSE
  )
} else {
  paddocks_sf <- NULL
  gps_spatial <- gps_dedup
}

```

Basic QA plot: 10% sample of points + paddock boundary.
Because filtering is done with a 100 m buffer, we show both:
- solid line = original boundary
- dashed line = +100 m buffered extent used for filtering

```{r}
set.seed(101)
plot_idx <- sample.int(nrow(gps_spatial), max(1L, floor(0.10 * nrow(gps_spatial))))
gps_plot <- gps_spatial[plot_idx, ]
paddock_buffer_sf <- NULL

if (has_sf && !is.null(paddocks_sf)) {
  paddock_union <- sf::st_union(sf::st_geometry(paddocks_sf))
  ctr <- sf::st_coordinates(sf::st_centroid(paddock_union))
  utm_zone <- floor((ctr[1] + 180) / 6) + 1
  utm_epsg <- if (ctr[2] >= 0) 32600 + utm_zone else 32700 + utm_zone

  paddock_buffer_sf <- paddocks_sf %>%
    sf::st_transform(utm_epsg) %>%
    sf::st_buffer(dist = 100) %>%
    sf::st_transform(4326)

  p <- ggplot() +
    geom_point(
      data = gps_plot,
      aes(x = lon, y = lat),
      alpha = 0.3,
      size = 0.8,
      color = "#2b8cbe"
    ) +
    theme_minimal() +
    labs(
      title = "10% GPS sample with paddock boundary (+100 m buffer)",
      x = "Longitude",
      y = "Latitude"
    )

  p <- p +
     geom_sf(data = paddock_buffer_sf, fill = NA, color = "black", linetype = "22", linewidth = 0.5, inherit.aes = FALSE) +
     geom_sf(data = paddocks_sf, fill = NA, color = "black", linewidth = 0.6, inherit.aes = FALSE)

  p
} else {
  ggplot(gps_plot, aes(x = lon, y = lat)) +
    geom_point(alpha = 0.3, size = 0.8, color = "#2b8cbe") +
    theme_minimal() +
    labs(
      title = "10% GPS sample (paddock boundary skipped: sf not available)",
      x = "Longitude",
      y = "Latitude"
    )
}
```

### 3c) Filter extreme speeds

This is a heuristic step. You must decide what is biologically realistic for
your herd/context.

In future work, we intend to provide more formal statistical guidance.

```{r}
mv_before_speed <- grz_calculate_movement(gps_spatial, verbose = FALSE)

gps_speed <- grz_clean_speed_fixed(
  data = gps_spatial,
  max_speed_mps = 3,
  keep_speed_cols = FALSE,
  verbose = FALSE
)

mv_after_speed <- grz_calculate_movement(gps_speed, verbose = FALSE)
```

```{r}
speed_threshold <- 3

spd_summary <- tibble(
  stage = c("Before speed filter", "After speed filter"),
  speed = list(mv_before_speed$speed_mps, mv_after_speed$speed_mps)
) %>%
  unnest(speed) %>%
  filter(is.finite(speed), speed >= 0) %>%
  mutate(over_threshold = speed > speed_threshold) %>%
  summarise(
    fixes_over_threshold = sum(over_threshold),
    total_fixes = n(),
    pct_over_threshold = 100 * fixes_over_threshold / total_fixes,
    .by = stage
  )

spd_summary
```

Why do we still have speeds over 3mps?

### 3d) Denoise

Most livestock GPS trackers are single-frequency (L1) devices. In practical
terms, that means location error and local jitter are expected, especially when
animals are stationary.

`grz_denoise()` supports:
- `method = "statistical"`: spline smoothing on all points.
- `method = "state_aware"`: collapse inactive runs to a robust median location,
  then spline-smooth active runs.

Depending what you're hoping to use it for, this step may not be required. Furthermore, `state_aware` denoise procedures will take some time to process.

#### 3d-i) Create ACTIVE/INACTIVE labels with GMM + HMM smoothing

We use movement features (mainly step and turning behaviour) to
separate "low movement" vs "higher movement" states. HMM smoothing reduces
point-to-point flicker in state labels.

```{r}
gps_for_state <- grz_calculate_movement(gps_speed, verbose = FALSE)

gps_state <- grz_classify_activity_gmm(
  data = gps_for_state,
  smoothing = "hmm",
  feature_set = "adaptive",
  verbose = FALSE
)
```

Output Active/Inactive periods for one animal around the:

```{r}
focal_id <- gps_state %>%
  count(sensor_id, name = "n_fixes", sort = TRUE) %>%
  slice(1) %>%
  pull(sensor_id)
focal <- gps_state %>% filter(sensor_id == focal_id)

mid_time <- as.POSIXct(stats::median(as.numeric(focal$datetime), na.rm = TRUE), origin = "1970-01-01", tz = "UTC")
mid_day <- as.Date(mid_time, tz = "UTC")
window_start <- as.POSIXct(mid_day - 2, tz = "UTC")
window_end <- as.POSIXct(mid_day + 2, tz = "UTC")

focal_two_days <- focal %>%
  filter(datetime >= window_start, datetime < window_end) %>%
  arrange(datetime)

ggplot(focal_two_days, aes(x = lon, y = lat)) +
  geom_path(alpha = 0.45, color = "grey50", linewidth = 0.5) +
  geom_point(aes(color = activity_state_gmm), size = 1) +
  theme_minimal() +
  labs(
    title = paste0("Two-day state-labelled track: ", focal_id),
    x = "Longitude",
    y = "Latitude",
    color = "State"
  )
```

#### 3d-ii) Apply denoise algorithm

```{r}
gps_denoised <- grz_denoise(
  data = gps_state,
  method = "state_aware",
  state_col = "activity_state_gmm",
  keep_raw_coords = TRUE,
  verbose = FALSE
)
```

And the resulting data will look like this:

```{r}
focal_id <- gps_denoised %>%
  count(sensor_id, name = "n_fixes", sort = TRUE) %>%
  slice(1) %>%
  pull(sensor_id)
focal <- gps_denoised %>% filter(sensor_id == focal_id)

mid_time <- as.POSIXct(stats::median(as.numeric(focal$datetime), na.rm = TRUE), origin = "1970-01-01", tz = "UTC")
mid_day <- as.Date(mid_time, tz = "UTC")
window_start <- as.POSIXct(mid_day - 2, tz = "UTC")
window_end <- as.POSIXct(mid_day + 2, tz = "UTC")

focal_two_days <- focal %>%
  filter(datetime >= window_start, datetime < window_end) %>%
  arrange(datetime)

ggplot(focal_two_days, aes(x = lon, y = lat)) +
  geom_path(alpha = 0.45, color = "grey50", linewidth = 0.5) +
  geom_point(aes(color = activity_state_gmm), size = 1) +
  theme_minimal() +
  labs(
    title = paste0("Two-day state-labelled track: ", focal_id),
    x = "Longitude",
    y = "Latitude",
    color = "State"
  )
```

Does fitting splines make biological sense for 15 minute data? Consider the points in the top-left corner.

Statistical-only mode is still available:

```{r, eval = FALSE}
gps_denoised_stat <- grz_denoise(
  data = gps_speed,
  method = "statistical",
  keep_raw_coords = TRUE
)
```

### QA checkpoint: weekly fixes per animal

Here, we'd always recommend a check of the data. Have inadvertantly removed lots of data for one animal. This is up to the investigator to determine!!!

This plote helps highlight low-performing devices. We intentionally keep this as an
investigator decision, not an automated exclusion rule.

```{r}
study_start <- min(gps_denoised$datetime, na.rm = TRUE)
qa_week <- gps_denoised %>%
  mutate(
    week_index = floor(as.numeric(difftime(datetime, study_start, units = "days")) / 7) + 1L
  ) %>%
  count(sensor_id, week_index, name = "n_fixes")

max_week <- max(qa_week$week_index, na.rm = TRUE)

ggplot(qa_week, aes(x = week_index, y = sensor_id, fill = n_fixes)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C") +
  scale_x_continuous(
    breaks = seq_len(max_week),
    labels = paste0("W", sprintf("%02d", seq_len(max_week)))
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Weekly fix counts by animal",
    x = "Study week",
    y = "Animal",
    fill = "Fixes"
  )
```

## 4) Movement metrics (row-level)

Distance between fixes (`step_m`) and turning (`turn_rad`) are produced by
`grz_calculate_movement()`.

Grouping controls independent trajectories. By default, it groups by
`deployment_id + sensor_id` when available (otherwise `sensor_id`).

```{r}
gps_move <- grz_calculate_movement(
  data = gps_denoised,
  groups = "sensor_id",
  verbose = FALSE
)

gps_move %>%
  select(
    sensor_id, datetime, step_m, turn_rad, speed_mps,
    any_of(c("deployment_id", "paddock"))
  ) %>%
  head()
```

Metadata columns are retained, so you can keep treatment/paddock context in
downstream summaries.

There are wrappers that run default summary pipelines (for example
`grz_calculate_epoch_metrics()`), but if you need to tune assumptions, use the
individual functions directly as shown here.

## 5) Movement summaries and diurnal plots

First, use the movement summary function:

```{r}
move_day <- grz_summarise_movement(
  data = gps_move,
  epoch = "day",
  groups = "sensor_id",
  verbose = FALSE
)

head(move_day)
```

Now build diurnal line plots for distance and turning:

```{r}
diurnal <- gps_move %>%
  mutate(hour = as.integer(format(datetime, "%H", tz = "UTC"))) %>%
  summarise(
    mean_step = mean(step_m, na.rm = TRUE),
    mean_turn = mean(turn_rad, na.rm = TRUE),
    .by = hour
  )


diurnal_long <- bind_rows(
  diurnal %>% transmute(hour, metric = "Step distance (m)", mean = mean_step),
  diurnal %>% transmute(hour, metric = "Turn angle (rad)", mean = mean_turn)
)

ggplot(diurnal_long, aes(x = hour, y = mean)) +
  geom_line(color = "#045a8d", linewidth = 0.9) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  scale_x_continuous(breaks = seq(0, 23, by = 3)) +
  theme_minimal() +
  labs(
    title = "Diurnal movement profile (mean)",
    x = "Hour (UTC)",
    y = NULL
  )
```

## 6) Social metrics

We focus on:
- nearest neighbour distance
- number of neighbours within 30 m

`grz_calculate_social()` aligns/interpolates to a regular interval before
pairwise calculations when `interpolate = TRUE`.

```{r}
social_herd_groups <- if ("deployment_id" %in% names(gps_move)) "deployment_id" else NULL

gps_social <- grz_calculate_social(
  data = gps_move,
  thresholds_m = 30,
  align_interval_mins = "base",
  align_max_gap_mins = NULL,
  interpolate = TRUE,
  herd_groups = social_herd_groups,
  verbose = FALSE
)

n30_col <- grep("^n_within_30m$", names(gps_social), value = TRUE)
if (length(n30_col) == 0L) {
  n30_col <- grep("^n_within_.*30.*m$", names(gps_social), value = TRUE)
}
if (length(n30_col) == 0L) {
  stop("Could not find `n_within_30m`-style column in social output.", call. = FALSE)
}

gps_social %>%
  select(sensor_id, datetime, nearest_neighbor_m, all_of(n30_col[1])) %>%
  head()
```

## 7) Social summaries (daily, first 4 weeks)

```{r}
social_day <- grz_summarise_social(
  data = gps_social,
  epoch = "day",
  groups = "sensor_id",
  verbose = FALSE
) %>%
  mutate(date = as.Date(epoch))

date_min <- min(social_day$date, na.rm = TRUE)
social_4wk <- social_day %>%
  filter(date < date_min + 28)

n30_mean_col <- grep("^mean_n_within_30m$", names(social_4wk), value = TRUE)
if (length(n30_mean_col) == 0L) {
  n30_mean_col <- grep("^mean_n_within_.*30.*m$", names(social_4wk), value = TRUE)
}
if (length(n30_mean_col) == 0L) {
  stop("Could not find mean n_within_30m summary column.", call. = FALSE)
}

social_plot <- bind_rows(
  social_4wk %>% transmute(sensor_id, date, feature = "Mean nearest neighbour (m)", value = mean_nn_m),
  social_4wk %>% transmute(sensor_id, date, feature = "Mean n within 30m", value = .data[[n30_mean_col[1]]])
) %>%
  filter(is.finite(value))

if (nrow(social_plot) == 0L) {
  stop("Social summary produced no finite values for plotting. Check interpolation and herd grouping.", call. = FALSE)
}

ggplot(social_plot, aes(x = date, y = value, color = sensor_id)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~feature, scales = "free_y", ncol = 1) +
  theme_minimal() +
  labs(
    title = "Daily social summaries (first 4 weeks)",
    x = "Date",
    y = NULL,
    color = "Animal"
  )
```

## 8) Spatial summaries: weekly size and overlap for one animal

```{r}
hr_week <- grz_calculate_home_range(
  data = gps_denoised,
  method = "kde",
  epoch = "week",
  groups = "sensor_id",
  levels = 95,
  progress = TRUE,
  verbose = FALSE
)

hr_change <- grz_summarise_home_range_change(
  home_range = hr_week,
  level = 95,
  compare = "consecutive",
  include_map_layers = FALSE,
  verbose = FALSE
)
```

```{r}
hr_metrics <- as_tibble(hr_week$metrics)
hr_changes <- as_tibble(hr_change$change_metrics)

focal_id <- sort(unique(hr_metrics$sensor_id))[1]
hr_one <- hr_metrics %>% filter(sensor_id == focal_id)
if ("sensor_id" %in% names(hr_changes)) {
  chg_one <- hr_changes %>% filter(sensor_id == focal_id)
} else {
  chg_one <- hr_changes
}

size_col <- grep("^hr95_ha$", names(hr_one), value = TRUE)
if (length(size_col) == 0L) {
  size_col <- grep("^hr.*95.*_ha$", names(hr_one), value = TRUE)
}
if (length(size_col) == 0L) {
  stop("Could not find weekly 95% home-range size column.", call. = FALSE)
}

ggplot(hr_one, aes(x = epoch, y = .data[[size_col[1]]], group = 1)) +
  geom_line(color = "#045a8d", linewidth = 0.9) +
  geom_point(color = "#045a8d", size = 2) +
  theme_minimal() +
  labs(
    title = paste0("Weekly home-range size (95%) for animal ", focal_id),
    x = "Week",
    y = "Area (ha)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
if (nrow(chg_one) > 0) {
  ggplot(chg_one, aes(x = epoch_to, y = overlap_from_pct, group = 1)) +
    geom_line(color = "#2ca25f", linewidth = 0.9) +
    geom_point(color = "#2ca25f", size = 2) +
    theme_minimal() +
    labs(
      title = paste0("Week-to-week overlap (from previous week) for animal ", focal_id),
      x = "Week",
      y = "Overlap from previous week (%)"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  message("Not enough weekly epochs for overlap plotting in this subset.")
}
```

## 8b) Helper visualisation functions

### Heatmap helper

```{r}
grz_plot_diurnal_metrics(
    data = gps_move,
    metrics = c("step_m", "turn_rad"),
    group_col = "sensor_id",
    agg_fun = "median",
    scale = "none"
  )
```

### Leaflet helpers (code only)

```{r, eval = FALSE}
# Playback map
m <- grz_playback_gps(
  data = gps_denoised,
  group = "sensor_id",
  align = TRUE,
  align_interval_mins = "base",
  tail_points = 20,
  show_points = TRUE,
  playback_steps = 1000,
  playback_duration_ms = 12000
)
```

## 9) Conclusion

This baseline pipeline gets you to analysis-ready GPS movement, social, and
spatial metrics while keeping key cleaning decisions transparent.

From here, you can extend into more advanced spatial and resource-use analyses
(for example POI/resource exposure, weekly overlap modelling, and treatment
comparisons) with a reproducible foundation.
